Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-09 paper-ieee-trans.tex
revision corrections

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by a various number of projects dedicated to software development
and e-Infrastructure creation, operation and management.
% the creation of software development and e-Infrastructure coordination dedicated projects. 
This work examines the evolution of the software engineering methodologies used in past and
present European Commission-funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, hightlighting the main challenges and barriers confronted
at the time, and how they were totally or partially overcomed. The base of knowledge collected
throughout these years, sustained by the advances in the area of software engineering,
definitely conformed a significant breakthrough in the reliability of software delivered in the
European research arena. Latest software development projects founded by the European 
Commission are good evidences of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.
\end{abstract}

@@

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
which supports an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One of the well-known examples is the Worldwide 
Large Hadron Collider (LHC) Computing Grid infrastructure, powered 
by the particle physics scientific community, built up to satisfy the computing needs
of processing, storing and analysing the huge quantity of data coming from the 
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
and ultimately by the Horizon 2020 programme \cite{h2020} -- are being held to coordinate and 
develop distributed solutions that provide a seamless and transparent access to those
computing and data resources, available for the researchers.

Major developments were done throughout the years, mainly focused on providing new
functionalities, leaving aside in the early times the adoption of software engineering methodologies.
The software delivered was then less tested, resulting in constant
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e-Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

Soon it become apparent that a balance was needed between new features addition
and the reliability of the software produced. As a consequence, a substantial part of the
software lifecycle would have to be devoted to deal with the quality of software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define the quality procedures, teams
organization and pilot testbeds.

By analyzing the European Commission (EC)-funded software projects in the last decade
we have observed a continuous increase of the prominence and robustness of
testing and validation procedures. One of many reasons is related to the
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and 
operating--system--level virtualization known as containerization). All of which
complemented by the emergence of automation and event-response tools.

In this paper, we introduce a set of EC-funded projects
that go from 2001 up to now, in which the authors of this summary were involved.
For each project, we describe the challenges arised, the methodologies used and
the achievements towards software reliability. The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the 
experiences from the recent {\sl INDIGO--DataCloud} project. Finally, Section \ref{sec:con} concludes.

@@

The EC projects considered in this text are listed chronologically in Table
\ref{tab:eup}. In most cases, each project paved the way of the subsequent projects,
following a logic evolution in which the quality and reliability of software have
increasingly had a more significant role. Experiences gathered during the 
course of each project support this fact and, as it is shown in Table \ref{tab:feat},
latest projects are the most committed in terms of software reliability awareness.
Table \ref{tab:feat} summarizes also the topics addressed in the following sections, highlighting both
the commonalities and specific features on software reliability.

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
had as the main goal to provide scientific communities, including physics, biology and 
earth sciences, with intensive computing and large-scale dataset analysis capabilites.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}. 
Grid was the emerging technology to be used in order to address these requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
based on Globus middleware services \cite{globus}. The project was organized in 12 work packages, from 
which 5 were devoted to software development coordinated by an 
Architecture Task Force, supervising the overall design and technical consistency 
of the developments.

Due to having no previous experience in such widely
collaborative projects, a big effort was spent trying to devise solutions for several
challenges \cite{datagrid}, namely the 1) communication overhead, as a
result of the large geographical separation of the parties involved in the
development tasks, 2) the evolution of the requirements coming from the user
communities (50 use cases from the three scientific area), and 3) the lack of a body 
of knowledge for academic software engineering.

The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, thus not
considered by the project, and as such the project's
development design suffered from the methods, tools, techniques and best
practices emerging from the new discipline of software engineering
\cite{agile}. Nonetheless, the project focused on the experiences and procedures
used by diverse collaborative open source software projects, such as Linux and Apache, trying to 
reach a higher maturity level \cite{cmm}.

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from more than 240 institutions in 45 countries to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements from the diverse scientific communities would need to develop a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

The {\sl gLite} middleware \cite{glite} was the ultimate
official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
re--engineering efforts to converge with {\sl LHC} Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software distributions. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, that issued more than 10 thousand bug fixes, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

The source code was available at a private, centralized version control system.
The code passed through a manual certification procedure at the time of the release. 
This procedure tried to improve the reliability of the software components by applying an
acceptance criteria plan at each stage of the software release process 
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
production. Starting with {\sl EGEE--II}, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the {\sl ETICS} \cite{etics} solution.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
arose as the continuation of the {\sl EGEE--III} project, with the mission of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National 
Grid Initiatives and interoperable with other Grids worldwide.
To accomplish this goal the project foreseed the 
maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
the {\sl EGI} Helpdesk \cite{ggus} or the Grid configuration database (GOCDB) \cite{gocdb}, and the 
management of the software provisioning process \cite{mario}, to validate and distribute the software that conformed
the production infrastructure. As a result of this process, a production--ready 
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution) was
maintained, comprised by roughly 250 software components supplied by diverse technology providers such 
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The 
QC defines a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under a follow--up
project, the {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this review. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper thus it
emerged from the lessons learned in the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above with the additional complexity of
being framed into large collaborative projects with an extensive list of partners.

In the following sections we will describe the three main pillars that sustain the project's strategy to seek reliability throughout the 
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} will provide guidelines to develop quality 
software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is constituted by a \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

The software quality policies were initially defined in the first deliverable document of the
project \cite{indigo-d31} and periodically reviewed to be endowed with applicable insights of 
software engineering practices, responses to evolving needs emanated from the user communities and 
feedback obtained from development teams. These policies cover the 1) identification
and description of the \emph{SQA requirements} that the produced software need 
to comply with, and the 2) \emph{quality metrics}, to monitor each software product's 
behaviour throughout the development, release and post-release stages.

@@

The SQA criteria is a collection of conventions and recommendations that pave the way for
the adequate development, timely deliver and reliable operation of the software components
produced. It emphasizes the quality requirements and best practices to be applied at the
development phase, such as style and testing compliance or human code reviews, to protect the
production source code versions. Besides, the criteria promotes the adoption of the software 
by providing minimum requirements of documentation content and stressing the usage of automated
solutions for software deployment.

\begin{itemize}
\item \textit{Code style}.
The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
project. Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standards} are initially recommended, however the development teams eventually
choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
over the {\sl INDIGO--DataCloud} software stack.

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major {\sl INDIGO--DataCloud} releases, denoting an incremental 
trend over the course of the project. As depicted in the Figure, 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

\item \textit{Integration testing}. Software components usually interact with other services while on
operation state. Integration testing deals with the interactions among coupled software components or
parts of a system that cooperates to achieve a given functionality. This type of testing might be 
complex and, based on the project's experience, tough to be tackled automatically. The aim is to 
guarantee the overall operation of the component with regards to the services it interfaces with,
whenever new functionalities are introduced.

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has 
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequance in terms of scope, objective fulfillment, 
documentation completeness, etc. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially 
important when the SQA testing requirements are fully automated. Secure code reviews are
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

\item \textit{Documentation}.
The SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. These requirements also 
promote the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code}, using a markup language, automatically rendered and
uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and 
human--reviewed, using the same workflow as the code does, validating the changes before being
updated into the production repository.

\item \textit{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self-documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. In this regard, {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\textit{Release metric} sources are mainly the online repository servers, namely the 
Linux package \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins 
server contains as well valuable release data as it is the service where the packages
and containers are being built before being uploaded to the online repositories. Release
KPIs primarly focus on the frequency and effiency -- mainly rollbacks -- statistics.

\textit{Maintenance and user support metrics} are key for continuously improving the response
to issues reported by external users. Feeback is collected from outside helpdesks, such as 
EGI's GGUS \cite{ggus}, and the GitHub Issues tracker.

@@

The DevOps methods emphasize on exercising SQA techniques to avoid infrastructure
disruption whenever new developments are deployed into production systems. The 
following DevOps approaches have been progressively adopted throughout the project's
lifetime, being one of its major outcomes in terms of sofware reliability.

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforces,
for any piece of software produced, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as 
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline so to have a per-change trend evolution of the common source code 
related metrics, such as SLOC or the cyclomatic complexity.

The practical implementation of the CI scenario leveraged on tightly integrated open
source tools such as GitHub \cite{github} -- as the online source code repository --,
Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs. 
The project defined a source code contribution workflow, based on GitHub Pull Requests 
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline report their exit status back to GitHub, preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects augments. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected at this phase.

@@

The experience gathered throughout the project with regard to the adoption of
different DevOps practices is not only useful and suitable for the software related
to the core services in the {\sl INDIGO--DataCloud} solution, but also applicable to the
development and distribution of the applications coming from the user communities.

@@

The novelty introduced in the pipeline above is the validation of the application.
Once the application is packaged as a Docker image, and subsequently uploaded
to the Dockerhub repository, it is instantiated in a new container to be validated.
The application is then executed and the results compared with a set of reference outputs.
Thus this pipeline implementation goes a step forward by testing the application
execution for the latest available Docker image in the catalogue.

>>>>

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e-Infrastructure creation, operation and management.
% the creation of software development and e-Infrastructure coordination dedicated projects. 
This work examines the evolution of the software engineering methodologies used in past and
present European Commission-funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The base of knowledge collected
throughout the years, sustained by the advances in the area of software engineering,
definitely conformed a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European 
Commission are good evidences of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.
\end{abstract}

@@

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
supporting an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One such example is the Worldwide 
Large Hadron Collider (LHC) Computing Grid infrastructure, motivated 
by the particle physics scientific community, build to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the 5th, 6th and 7th Framework Programmes,
and recently by the Horizon 2020 programme \cite{h2020} -- are under way to
develop software solutions that provide a seamless and transparent access to distributed
computing and data resources by the researchers.

Major developments were achieved throughout the years focused on providing new
functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
The released software was less tested, resulting in a high rate of
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e-Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

Soon it become apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, a substantial part of the
software lifecycle would have to be devoted to the quality of the software being
delivered. The evolution of software engineering practices were
considered and gradually adopted to define and implement quality procedures, the organization of teams
and deployment of pilot testbeds.

Analysing the European Commission (EC)-funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures. One of many reasons, is related to the
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and 
operating--system--level virtualization known as containerization). All of which
complemented by the emergence of automation and event-response tools.

In this paper, we outline a set of EC-funded projects
ranging from 2001 to now, in which the authors were involved.
For each project, the challenges, the methodologies and
the achievements towards software reliability are described.

The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the 
experiences from the recent {\sl INDIGO--DataCloud} project. Conclusions are drawn in Section \ref{sec:con}.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way of the subsequent projects,
following a logic evolution in which the role of the quality and reliability of the software have
consistently increased. The experience gained during the 
course of each project support this fact as shown in Table \ref{tab:feat}.
The most recent projects are more committed and aware to software reliability practices.
Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
main goal was to provide scientific communities, including physics, biology and 
earth sciences, with intensive computing and large-scale dataset analysis capabilities.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}. 
Grid was the emerging technology to be used in order to address these requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from 
which 5 were devoted to software development and coordinated by an 
Architecture Task Force, supervising the overall design and technical consistency 
of the developments.

Without previous experience in such widely
collaborative projects, a big effort was spent in devising solutions for several
challenges \cite{datagrid}, namely the 1) communication overhead, as a
result of the large geographical separation of the involved parties in the
development tasks, 2) the fast evolution of the requirements from the user
communities (50 use cases from the three scientific area), and 3) the lack of a body 
of knowledge for academic software engineering.

The Agile manifesto \cite{agile-manifesto} was by then an emerging methodology, thus not
considered by the project, and as such the project's
development design suffered from the methods, tools, techniques and best
practices emerging from the new discipline of software engineering
\cite{agile}. Nevertheless, the project focused on the experiences and procedures
used by diverse collaborative open source software projects, such as Linux and Apache, trying to 
achieve a higher maturity level \cite{cmm}.

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements from the diverse scientific communities would need to develop a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

The {\sl gLite} middleware \cite{glite} was the
official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
re--engineering efforts to converge with the {\sl LHC} Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software stacks. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, fixing more than 10 thousand bug, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

The source code was available at a private, centralized version control system.
The code passed through a manual certification procedure at the time of the release. 
This procedure aimed to improve the reliability of the software components by applying an
acceptance criteria plan at each stage of the software release process 
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
production. Starting with {\sl EGEE--II}, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the {\sl ETICS} \cite{etics} solution.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
was the continuation of the {\sl EGEE--III} project, with the objective of establishing and
maintaining a sustainable European Grid Infrastructure, composed by a federation of National 
Grid Initiatives and interoperable with other Grids worldwide.
This goal would be accomplished through the
development and maintenance of operational tools, such as the Operations Portal \cite{egi-ops}, 
the {\sl EGI} Helpdesk \cite{ggus}, or the Grid configuration database (GOCDB) \cite{gocdb}, and the 
management of the software provisioning process \cite{mario}, to validate and distribute the software to
the production infrastructure. As a result, this process led to a production--ready 
Grid computing middleware distribution named {\sl UMD} (Unified Middleware Distribution). The {\sl UMD}
was a stack of about 250 software components from several technology providers such 
as {\sl EMI} and {\sl Globus}. In order to be distributed through UMD
repositories, the software has to be compliant with EGI's Quality Criteria (QC) \cite{egi-qc}. The 
QC defines a set of requirements in different areas: documentation, deployment, security, information
model and operations.
{\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under the follow--up
project {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this paper. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper thus, it
emerged from the lessons learned in the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above with the additional complexity of
large collaborative projects with an extensive list of partners.

In the following sections, we will describe the three main pillars that sustain the project's strategy to attain reliability throughout the 
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition}, will provide guidelines to develop quality 
software, validating each change in the code, and facilitate its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} will promote automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

The software quality policies were initially defined in the first deliverable document of the
project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
from the user communities and feedback from the development teams. These policies cover the 1) identification
and description of the \emph{SQA requirements} that the produced software need 
to comply with, and the 2) \emph{quality metrics}, to monitor each software product's 
behaviour throughout the development, release and post--release stages.

@@

The SQA criteria is a set of conventions and recommendations that pave the way for
the adequate development, timely deliver and reliable operation of the produced software components.
It emphasizes the quality requirements and best practices to be applied at the
development phase, such as style and testing compliance or human code reviews, to protect the
production source code versions. Furthermore, the criteria promotes the adoption of the software 
by providing minimum requirements of documentation content and stressing the usage of automated
solutions for software deployment.

\begin{itemize}
\item \textit{Code style}.
The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
project. Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standards} are initially recommended, however the development teams eventually
choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
in the {\sl INDIGO--DataCloud} software stack.

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities, are required to be tested. \textit{Regression
testing} in this context is accomplished by enforcing the successful execution of the previous
functionality tests already present in the code to uncover any malfunction introduced by
the changes. \textit{Unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental 
trend over the course of the project. It can be seen that 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

\item \textit{Integration testing}. Software components usually interact with other services during
operation. Integration testing deals with the interactions among coupled software components or
parts of a system that cooperate to achieve a given functionality. This type of testing might be 
complex and, based on the project's experience, difficult to be implemented in an automatic way. The aim is to 
guarantee the overall operation of the component with regard to the services it interfaces with,
whenever new functionalities are introduced.

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has 
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequacy in terms of scope, objective fulfilment, 
documentation completeness, etc. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially 
important when the SQA testing requirements are fully automated. Secure code reviews are
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

\item \textit{Documentation}.
The SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. This requirement also 
promotes the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code}, using a markup language, automatically rendered and
uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and 
human--reviewed, using the same workflow as the code does, validating the changes before being
updated into the production repository.

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self-documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\textit{Release metric} sources are the online repository servers, namely the 
Linux package \cite{indigo-pkg-repo} and DockerHub \cite{indigo-dockerhub} repositories. Jenkins 
server also contains valuable release data as it is the service where the packages
and containers are being built before being uploaded to the online repositories. Release
KPIs primarily focus on the frequency and efficiency -- mainly rollbacks -- statistics.

\textit{Maintenance and user support metrics} are key for continuously improving the response
to issues reported by external users. Feedback is collected from outside helpdesks, such as 
EGI's GGUS \cite{ggus}, and the GitHub Issues tracker.

@@

The DevOps methods emphasize the SQA techniques to avoid infrastructure
disruption whenever new developments are deployed into production systems. The 
following DevOps approaches have been progressively adopted throughout the project's
lifetime, being one of its major outcomes in terms of software reliability.

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforces,
for any piece of produced software, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as 
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline to have a per-change trend evolution of the common source code 
related metrics, such as SLOC or the cyclomatic complexity.

The practical implementation of the CI scenario leveraged on tightly integrated open
source tools such as GitHub \cite{github} -- as the online source code repository --,
Jenkins \cite{jenkins} -- the event-response CI system -- and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs. 
The project defined a source code contribution workflow, based on GitHub Pull Requests 
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

The experience gathered throughout the project regarding the adoption of
different DevOps practices is not only useful and suitable for the software related
to the core services in the {\sl INDIGO--DataCloud} solution, but also applicable to the
development and distribution of the applications coming from the user communities.

@@

The novelty introduced in the pipeline above is the validation of the application.
Once the application is packaged as a Docker image, and subsequently uploaded
to the Dockerhub repository, it is instantiated in a new container to be validated.
The application is then executed and the results compared with a set of reference outputs.
This pipeline implementation goes a step forward by testing the application
execution for the latest available Docker image in the catalogue.
